---
title: "Semantic"
author: "Wolfgang Kaisers"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
%\VignetteIndexEntry{Semantic}
%\VignetteEngine{knitr::rmarkdown}
\usepackage[utf8]{inputenc}
---

```{r, eval=FALSE}

```

# Header settings

# The RDF data model

The RDF data model is described in section 12 of
[SPARQL Query Language for RDF](http://www.w3.org/TR/rdf-sparql-query/).

## RDF Term
An RDF term as defined in 
[SPARQL Query Language](http://www.w3.org/TR/rdf-sparql-query/#defn_RDFTerm)
is either
 - an IRI
 - an RDF literal or
 - a blank node.


### IRI's

IRIs are a generalization of URIs. It is common practice to abbreviate
IRI's by prefixes.

Prefixes are declared by the PREFIX keyword: 
The PREFIX keyword associates a prefix label with an IRI.

Inside an IRI the colon ":" separates the prefix name from the local part.


IRI's are define in the 
[RDF Concepts and abstract syntax](https://dvcs.w3.org/hg/rdf/raw-file/default/rdf-concepts/index.html#section-IRIs)
document as string denoting a resource. 
An IRI must not contain one of the characters *<>" {}|\1`*

 - International Resource Identifier (IRI)
 - Generalization of URI: allows non-ASCII characters
 - IRIs are specified in [RFC 3987](http://www.ietf.org/rfc/rfc3987.txt)
 - Full IRIs are enclosed in angle brackets (<>)
 - The datatype is appended to the literal through a ^^ delimiter.
 
Examples for IRI's are 
```c
<http://example.org/book/book1>
```
or
```c
BASE <http://example.org/book/>
<book1>
```
or
```c
PREFIX book: <http://example.org/book/>
book:book1
```

### RDF Literals

The general syntax for literals is a string (possibly enclosed b single or
double quotes: " ').
RDF Literals may contain a *language tag* (e.g. @en) or a *datatype URI*.

Examples for literals are:
```c
"cat"
"cat"@en
"10"^^xsd:integer
```

### Blank nodes

Blank node labels (section 4.1.4) may be written as "_:abc" for a blank node
with label "abc" or  as "[]" (abbreviation for singular use).
A blank node can only be used inside a single graph pattern in the
same query.

Use cases for blank nodes are 
```c
_:abc :p ?o .
[ :p ?o ] .
[] :p ?o .
```


## Triple patterns

A Triple pattern as described in 12.1.4

RDF triples 
```c
<subject> <predicate> <object>
```
are defined as element of
```c
(RDF-T union V) x (I union V) x (RDF-T union V)
```
where RDF-T denotes an RDF-term and V a query variable.


 - Subject and the object represent the two resources being related;
 - The predicate represents the nature of their relationship. 
 - The relationship is called in RDF a property. 
 - Because RDF statements consist of three elements they are called triples. 


## Graph patterns

### Basic Graph Patterns

A basic graph pattern is a set of Triple Patterns.
Basic graph patterns are grouped by braces "{}".
An example is given by where the same blank node "_:abc" is linked to 
two different query variables ("?o1", ("?o2") via two predicates (:p1 and :p2):
```c
{
    _:abc :p1  ?o1 .
    _:abc :p2  ?o2 .
}
```


### Group Graph Pattern

Multiple Basic Graph Patterns can be grouped together:

```c
{
    { ?x foaf:name ?name . }
    { ?x foaf:mbox ?mbox . }
}
```

The empty group pattern "{}" can also be used.
It is defined to match any graph.




## The SPARQL protocol

[sparql11-protocol](http://www.w3.org/TR/2013/REC-sparql11-protocol-20130321/)
desccribes the SPARQL Protocol. Section 1.2 Terminology describes

 - *SPARQL Protocol client* (client) as an HTTP client that sends HTTP requests
    for SPARQL protocol operations.
 - *SPARQL Protocol service* (service) as an HTTP server that services 
    HTTP requests and sends back HTTP responses for SPARQL Protocol operations.
    The URI at which a SPARQL Protocol service listens for requests
    is generally known as a SPARQL endpoint. 

### SPARQL protocol operations ###

SPARQL defines to operation types: query and update. Operations and 
returned messages are sent via HTTP.

 - The HTTP method by which the request is sent.
 - The HTTP query string parameters included in the HTTP request URI.
 - The message content included in the HTTP request body.
 - The message content included in the HTTP response body.

### Response formats

Response format should be selected by HTTP content negotiation.

SPARQL query answers (to SELECT and ASK) can be given as

 - XML
 - JSON
 - CSV/TSV

SPARQL query answers (to DESCRIBE and CONSTRUCT) return an RDF graph

## SPARQL syntax

### Prefix declaration

SPARQL queries allow the user to abbreviate URIs by declaring prefix
abbreviations. 
For example, the declaration
``` c
PREFIX dcterms: <http://purl.org/dc/terms/>
```

allows the URI *http://purl.org/dc/terms/Location* to be abbreviated as 
*dcterms:Location*. [Prefixes](http://prefix.cc/)

SPARQL bases on matching parts of RDF graphs: The pattern
``` c
?Location  dwc:stateProvince "Hawaii"
```
matches
```c
<http://bioimages.vanderbilt.edu/baskauf/04005#loc>  dwc:stateProvince  "Hawaii"
```
A triple pattern can also contain zero, one, two, or three variables.

# The S4 class structure inside semantic

## Namespace objects

Namespace (PREFIX) data is managed in class *Namespace*. Therefore functions 
for imporing, subspacing and export of Namespace data are provided.

From an existing *Namespace* object, the raw textual representation can
be obtained using (cat(asPrefix(nx))) as shown below.
 
```{r}
library(semantic)
ns <- namespace("PREFIX owl: <http://www.w3.org/2002/07/owl#>")
cat(asPrefix(ns))
```

Example namespaces:

 - http://www.w3.org/TR/swbp-xsch-datatypes/ : 1.2 Namespaces used in this document


# Applications

We provide example use cases for the *semantic* package.

## Query EBI databases

Example SPARQL Queries are shown
[here](http://www.ebi.ac.uk/rdf/example-sparql-queries).



```{r}
ef <- system.file("extdata", "ebi_ns.txt", package="semantic")
ns <- importNamespace(ef)
```

### Query Chembl

First a *Sparql* object is created and the connection to
the chembl SPARQL endpoint is tested.
```{r, eval=FALSE}
ep <- sparql("http://www.ebi.ac.uk/rdf/services/chembl/sparql")
validObject(ep)
```

We query the chembl identifier for Temozolomide.
```{r, eval=FALSE}
en <- ns["skos"]
sql <-"SELECT ?mol WHERE { ?mol skos:altLabel 'Temozolomide' . }" 
sp <- sparqlQuery(sql, en)
res <- sendQuery(ep, sp, accept="table")
rs <- resultSet(res)
```
The query returns one record. Therefrom the core information
can be retrieved as data.frame.
```{r, eval=FALSE}
dfr <- as.data.frame(rs)
```


EBI RDF platform opened on 8 Oct 2013

 - [ChEMBL-og](http://chembl.blogspot.de/2013/10/embl-ebi-rdf-platform.html)
 - [EBI documentation](https://www.ebi.ac.uk/rdf/documentation/sparql-endpoints)
    Provides information on pre-defined PREFIXES.
 - [EBI Infrastructure](http://www.ebi.ac.uk/rdf/documentation/infrastructure)
 - [Archiv Chem2Bio2RDF](http://arxiv.org/pdf/1012.4759.pdf)
    (= PMID 20478034)

 - IUPAC Pharmacology [@pmid24234439]
 - ChEMBL linked open data [@pmid23657106]
 - Semantic linked data [@pmid22859915]
 - Bio LDA [@pmid21448266]
 - BioRDF [@pmid23735196]
 
 
# SPARQL Server

## Apache Jena Fuseki

 - Download binary from 
[Apache](http://mirror.arcor-online.net/www.apache.org//jena/binaries/jena-fuseki-1.1.1-distribution.tar.gz).
 - Unzip
 - Grant permission to execute server: chmod +x fuseki-server s-*
 - Run server: ./fuseki-server --update --mem /ds (on port 3030)
 - Open page [fuseki](http://localhost:3030/) page on local machine.
 - See documentation on [apache](http://jena.apache.org/documentation/serving_data/).
 - [Documentation](http://jena.apache.org/documentation/)

<!-- https://baach.de/Members/jhb/getting-started-with-jena-fuseki-and-owl-reasoning --!>

Fuseki can use a (nonpersistent) RAM database, a tdb database and 
mysql (no OWL support on mysql).

See also [FaCT](https://code.google.com/p/factplusplus/):
FaCT++ is a DL reasoner. It supports OWL DL and (partially) OWL 2 DL.

```c
./fuseki-server --update --loc ~/projects/java/apache_jena/jena-fuseki/db /ds
```

### Install 2

 - [Download](http://filteredpush.sourceforge.net/pages/fuseki.html)


```c
cd Downloads
tar -xzvf jena-fuseki-1.1.1-distribution.tar.gz
sudo mkdir /usr/share/fuseki
sudo mv jena-fuseki-1.1.1 /usr/share/fuseki
```

Download and install assembler config
```c
wget http://sourceforge.net/projects/filteredpush/files/Release_1/misc/tdb-assembler.ttl 
sudo cp tdb-assembler.ttl /usr/share/fuseki/jena-fuseki-1.1.1/
```


## Bigdata 


Bigdata can be downloaded an started via
```c
cd /path/to/bigdata/installation
git clone git://git.code.sf.net/p/bigdata/git bigdata-git
cd bigdata-git
ant start-bigdata
```
(see [bigdata download](http://www.bigdata.com/download)).
The bigdata user interface can be opened on a browser on [port 9999](http://localhost:9999/bigdata/#query).






 
 

# Bibliography



